A formalization of semi-simplicial and semi-cubical sets in dependent type theory

### 2023-01-11 paper/paper.tex
art.cls, paper: mostly whitespace changes, update submodule

This paper describes an alternative indexed construction of both augmented semi-simplicial and semi-cubical sets based on the property that augmented semi-simplicial and semi-cubical sets naturally arise as respectively unary and binary iterated parametricity-based constructions.

@@

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds72} interprets types as relations characterizing the observational equality of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has a cubical flavor~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates instead of relations, in which case we obtain a form of realizability~\cite{bernardy12,moulin16}. It has then been noted that iterated unary parametricity has an augmented simplicial flavor~\footnote{Private communication by Thorsten Altenkirch and Hugo Moeneclaey}.

We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both semi-simplicial and semi-cubical sets obtained by applying simple rules iterating parametricity. In contrast to the standard presheaf presentation of semi-simplicial and semi-cubical sets~\cite{fri08,grandis03,buchholtz1}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of semi-simplices or semi-cubes in dimension $n+1$
\emph{fibered} over the set of semi-simplices or semi-cubes in dimension $n$, we consider families of semi-simplices and semi-cubes indexed over their faces.

@@

\subsection{Semi-cubical sets}

Semi-cubical sets are defined like augmented semi-simplicial sets except that $\DeltaPlus$ is replaced by $\Cube$ in which we take sequences of $L$, $R$ and $\star$, instead of sequences of $0$ and $\kstar$.

@@

\section{Our setting}

In this

\subsection{Working with sets in type theory}
Martin-L\"of's Type theory~\cite{martinlof84,?} is a logical formalism based on the notion of $\Type$ rather than \Set. It is the core of several tools for the formalization of mathematics such as Coq~\cite{} or Lean~\cite{}. Type Theory is a flexible formalism supporting different models. One of this model interprets types as topological spaces, where equality is interpreted as path, and substitutivity of equality as transport, leading to the development of Homotopy Type Theory~\cite{hottbook}, a foundation of mathematics alternative to set theory. Conversely, a notion of sets can be recovered within Type Theory as \U, denoting the subtype of $\Type$ for which paths are degenerated, what can be expressed by the property of Uniqueness of Identity Proofs (UIP). Technically, this is expressed as a structure equipping a domain $\Dom$ with the property $\UIP$:

@@

\begin{enumerate}
  \item[(i)] UIP holds on the unit and bool type:
\begin{align*}
  \textsf{eq\_refl}        & : \forall x, x = x                                                    \\
  \textsf{UIP\_refl\_unit} & : \forall u : \textsf{unit}, \forall x : u = u, x = \textsf{eq\_refl} \\
  \textsf{UIP\_refl\_bool} & : \forall b : \textsf{bool}, \forall x : b = b, x = \textsf{eq\_refl} \\
\end{align*}
  \item[(ii)] UIP propagates to $\Sigma$-types.
  \item[(iii)] UIP propagates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

@@

\section{Our construction}

In this section, we gradually build up an intuition for our construction, followed by the full details.

\subsection{Intuition for our formal construction}

\begin{notation}[$\Sigma$]
  In the prose that follows, we assume the existence of $\Sigma$-types with projections $\hd$ and $\tl$.
\end{notation}

>>>>

  This paper describes an alternative indexed construction of both augmented semi-simplicial and semi-cubical sets based on the property that they naturally arise as respectively unary and binary iterated parametricity-based constructions.

@@

\section{Introduction}
In the context of functional programming, Reynolds' parametricity~\cite{reynolds72} interprets types as relations characterizing the observational equality of programs of this type. Parametricity can be iterated, and it has been noted that iterated Reynolds' parametricity has a cubical flavor~\cite{johann17,altenkirch15,moulin16,moeneclaey21,moeneclaey22phd}. We obtain a \emph{unary} variant of Reynolds' \emph{binary} parametricity by using predicates instead of relations, in which case we obtain a form of realizability~\cite{bernardy12,moulin16}. It has then been noted that iterated unary parametricity has an augmented simplicial flavor~\footnote{Private communication with Thorsten Altenkirch and Hugo Moeneclaey}.

We exploit the connection between iterated unary parametricity and augmented simplicial sets, and between iterated binary parametricity and cubical sets to present a uniform construction of both semi-simplicial and semi-cubical sets obtained by applying simple rules iterating parametricity. In contrast to the standard presheaf presentation of semi-simplicial and semi-cubical sets~\cite{fri08,grandis03,buchholtz17}, our uniform construction, which we call \emph{$\nu$-sets}, is \emph{indexed}. That is, instead of having the set of semi-simplices or semi-cubes in dimension $n+1$ \emph{fibered} over the set of semi-simplices or semi-cubes in dimension $n$, we consider families of semi-simplices and semi-cubes indexed over their faces.

@@

\subsection{Semi-cubical sets}
Semi-cubical sets are defined like augmented semi-simplicial sets except that $\DeltaPlus$ is replaced by $\Cube$ in which we take sequences of $L$, $R$ and $\star$, instead of sequences of $0$ and $\kstar$.

@@

\section{Our setting}
In this section,

\subsection{Working with sets in type theory}
Martin-Löf's Type theory~\cite{martinlof84,?} is a logical formalism based on the notion of $\Type$ rather than \Set. It is the core of several tools for the formalization of mathematics such as Coq~\cite{} or Lean~\cite{}. Type Theory is a flexible formalism supporting different models. One of this model interprets types as topological spaces, where equality is interpreted as path, and substitutivity of equality as transport, leading to the development of Homotopy Type Theory~\cite{hottbook}, a foundation of mathematics alternative to set theory. Conversely, a notion of sets can be recovered within Type Theory as \U, denoting the subtype of $\Type$ for which paths are degenerated, what can be expressed by the property of Uniqueness of Identity Proofs (UIP). Technically, this is expressed as a structure equipping a domain $\Dom$ with the property $\UIP$:

@@

\begin{enumerate}
  \item[(i)] UIP holds on the unit and bool type:

    \begin{align*}
      \textsf{eq\_refl}        & : \forall x, x = x                                                    \\
      \textsf{UIP\_refl\_unit} & : \forall u : \textsf{unit}, \forall x : u = u, x = \textsf{eq\_refl} \\
      \textsf{UIP\_refl\_bool} & : \forall b : \textsf{bool}, \forall x : b = b, x = \textsf{eq\_refl} \\
    \end{align*}

  \item[(ii)] UIP propagates to $\Sigma$-types.
  \item[(iii)] UIP propagates to $\Pi$-types, with some additional functional extensionality axioms.
\end{enumerate}

@@

\section{Our construction}
In this section, we gradually build up an intuition for our construction, followed by the full details.

\subsection{Intuition for our formal construction}
\begin{notation}[$\Sigma$]
  In the prose that follows, we assume the existence of $\Sigma$-types with projections $\hd$ and $\tl$.
\end{notation}
